Measuring matter effects in Neutron star - Black hole binaries with Advanced LIGO

### 2016-04-01 Paper/paper.tex
add snr threshold plots for single obs
S3 written as first draft

% #################
\begin{figure*}
\centering 
%\includegraphics[trim={11cm 0 0 0},width=2.7\columnwidth]{plots/TNMchirpBiasesOverCIWidths_CI68_3_Lambda_SNR.pdf}\\
% \includegraphics[trim={3.6cm 0 0 0},width=2.25\columnwidth]{plots/TNMchirpBiasesOverCIWidths_CI90_0_Lambda_SNR}
\includegraphics[trim={1.3cm 0 0 0},width=2.\columnwidth]{plots/TNMchirpBiasesOverCIWidths_CI90_0_Lambda_SNR30_70_linear}
\caption{We show here the ratio of systematic and statistical
measurement uncertainties for the binary chirp mass over the NSBH parameter 
space. Each panel shows the same as a function of BH mass and spin. Across
each row, we can see the effect of increasing the signal strength (i.e. SNR)
with the NS's deformability fixed. Down each column, we can see 
the effect of the increasing NS's tidal deformability at fixed SNR. We also
show dashed contours for where the ratio equals $10\%$ (labelled ``-1'')
or $100\%$ (labelled ``0'').
% 
Generally, these two sources of uncertainty are different and for BBHs, the
statistical errors dominate systematic ones~\cite{Kumar:2016dhh}. We find
that for BHNS binaries, its not much different until we get to high SNRs
$\rho\gtrsim 70$ for which if the BH is spinning rapidly enough, the 
measurement uncertainty for $\mchirp$ could become dominated by the exclusion
of tidal effects in filtering templates.
}
\label{fig:TN_chirpMassBias_vs_Lambda_SNR}
\end{figure*}
%
\begin{figure*}[!t]
\centering    
%\includegraphics[trim={11cm 0 0 0},width=2.7\columnwidth]{plots/TNEtaBiasesOverCIWidths_CI68_3_Lambda_SNR.pdf}\\
% \includegraphics[trim={3.6cm 0 0 0},width=2.25\columnwidth]{plots/TNEtaBiasesOverCIWidths_CI90_0_Lambda_SNR.pdf}\\
\includegraphics[trim={1.3cm 0 0 0},width=2.\columnwidth]{plots/TNEtaBiasesOverCIWidths_CI90_0_Lambda_SNR30_70_linear002}
\caption{This figure is similar to Fig.~\ref{fig:TN_chirpMassBias_vs_Lambda_SNR}
with the difference that the quantity in question here is the symmetric 
mass-ratio $\eta$. We find that for fairly loud GW signals, with $\rho\simeq 50$,
not including the effects of NS's tidal deformation on GW emission can become the dominant
source of error for astrophysical searches with Advanced LIGO. However,
for quieter signals with $\rho\leq 30$, it will have minimal effects on the measurement
of $\eta$. We remind the reader that the SNRs here are all single detector values.
}
\label{fig:TN_EtaBias_vs_Lambda_SNR}
\end{figure*}
% 
% 
\begin{figure*}
\centering
\includegraphics[trim={1.3cm 0 0 0},width=2.\columnwidth]{plots/TNChiBHBiasesOverCIWidths_CI90_0_Lambda_SNR_linear}
% \includegraphics[width=\columnwidth]{plots/TNChiBHBiasesOverCIWidthsVsSNR_All_CI68_3.pdf}
% \includegraphics[width=\columnwidth]{plots/TNChiBHBiasesOverCIWidthsVsSNR_All_CI90_0.pdf}
\caption{This figure shows the ratio of the systematic and statistical
measurement errors for BH spins, with other attributed identical to 
Fig.~\ref{fig:TN_chirpMassBias_vs_Lambda_SNR}, and~\ref{fig:TN_EtaBias_vs_Lambda_SNR}.
Similar to the case of mass parameters, we find that below
$\rho\approx 30$, ignoring tidal effects in templates introduces minor systematic effects,
which remain subdominant to the statistical measurement uncertainties.
}
\label{fig:TNT_BHspinBias_vs_SNR_q23}
\end{figure*}
% 
% 
Advanced LIGO searches and parameter estimation efforts aimed at binaries 
containing NS and stellar-mass BHs are poised to use waveform models that do not
include the effects of the NS tidal deformability~\cite{Canton:2014ena}.
While this is not expected to
be the dominant source of error at low SNRs, it will likely introduce a
systematic bias in the physical parameters that we recover from GW observations.
In this section, we study the impact of the same for binaries where the BH spins
are aligned to the orbital angular momentum.

\begin{enumerate}
\item Above what SNR values, below what mass-ratios, above what BH spins, etc, 
do we begin to care about NS deformability?\newline
``one interesting plot to make would be the parameter bias (max-Likelihood
parameters minus true parameters) vs SNR, as functionals of q/sBH/Lambda, when
using T signals recovered with N templates.''\newline
\end{enumerate}

\hspace{5mm}

@@

>>>>

% #################
%\includegraphics[trim={11cm 0 0 0},width=2.7\columnwidth]{plots/TNMchirpBiasesOverCIWidths_CI68_3_Lambda_SNR.pdf}\\
% \includegraphics[trim={3.6cm 0 0 0},width=2.25\columnwidth]{plots/TNMchirpBiasesOverCIWidths_CI90_0_Lambda_SNR}
\includegraphics[trim={1.3cm 0 0 0},width=2.\columnwidth]{plots/TNMchirpBiasesOverCIWidths_CI90_0_Lambda_SNR30_70_linear}
\caption{We show here the ratio of systematic and statistical
measurement uncertainties for the binary chirp mass over the NSBH parameter 
space. Each panel shows the same as a function of BH mass and spin. Across
each row, we can see the effect of increasing the signal strength (i.e. SNR)
with the NS's deformability fixed. Down each column, we can see 
the effect of the increasing NS's tidal deformability at fixed SNR. We also
show dashed contours for where the ratio equals $10\%$ (labelled ``-1'')
or $100\%$ (labelled ``0'').
Generally, these two sources of uncertainty are different and for BBHs, the
statistical errors dominate systematic ones~\cite{Kumar:2016dhh}. We find
that for BHNS binaries, its not much different until we get to high SNRs
$\rho\gtrsim 70$ for which if the BH is spinning rapidly enough, the 
measurement uncertainty for $\mchirp$ could become dominated by the exclusion
of tidal effects in filtering templates.
\label{fig:TN_chirpMassBias_vs_Lambda_SNR}
\begin{figure*}[!t]
%\includegraphics[trim={11cm 0 0 0},width=2.7\columnwidth]{plots/TNEtaBiasesOverCIWidths_CI68_3_Lambda_SNR.pdf}\\
% \includegraphics[trim={3.6cm 0 0 0},width=2.25\columnwidth]{plots/TNEtaBiasesOverCIWidths_CI90_0_Lambda_SNR.pdf}\\
\includegraphics[trim={1.3cm 0 0 0},width=2.\columnwidth]{plots/TNEtaBiasesOverCIWidths_CI90_0_Lambda_SNR30_70_linear002}
\caption{This figure is similar to Fig.~\ref{fig:TN_chirpMassBias_vs_Lambda_SNR}
with the difference that the quantity in question here is the symmetric 
mass-ratio $\eta$. We find that for fairly loud GW signals, with $\rho\simeq 50$,
not including the effects of NS's tidal deformation on GW emission can become the dominant
source of error for astrophysical searches with Advanced LIGO. However,
for quieter signals with $\rho\leq 30$, it will have minimal effects on the measurement
of $\eta$. We remind the reader that the SNRs here are all single detector values.
\label{fig:TN_EtaBias_vs_Lambda_SNR}
\includegraphics[trim={1.3cm 0 0 0},width=2.\columnwidth]{plots/TNChiBHBiasesOverCIWidths_CI90_0_Lambda_SNR_linear}
% \includegraphics[width=\columnwidth]{plots/TNChiBHBiasesOverCIWidthsVsSNR_All_CI68_3.pdf}
% \includegraphics[width=\columnwidth]{plots/TNChiBHBiasesOverCIWidthsVsSNR_All_CI90_0.pdf}
\caption{This figure shows the ratio of the systematic and statistical
measurement errors for BH spins, with other attributed identical to 
Fig.~\ref{fig:TN_chirpMassBias_vs_Lambda_SNR}, and~\ref{fig:TN_EtaBias_vs_Lambda_SNR}.
Similar to the case of mass parameters, we find that below
$\rho\approx 30$, ignoring tidal effects in templates introduces minor systematic effects,
which remain subdominant to the statistical measurement uncertainties.
\label{fig:TN_BHspinBias_vs_Lambda_SNR}
Past and future GW search and source characterization efforts with Advanced LIGO
have used (or plan to) BH-BH waveform templates for NS-BH mergers, ignoring
the effect of the tidal deformation of the neutron star due to its companion hole
on the emitted GWs. In this section we present the first fully Bayesian Monte-Carlo
study of the effects of this assumption on the recovery of other physical parameters
of NS-BH sources. We do so by comparing the systematic biases introduced by this 
assumption to the overall statistical uncertainties in the measurement of binary 
mass and spin parameters. We find that unless the single-detector SNR of the signal
exceeds $30$ (for design Advanced LIGO sensitivity), the bias accurued by ignoring
tidal effects in filtering templates will remain smaller than the statistical 
uncertainty of the same measurement.

We first take a set of NSBH binary waveforms, which include the modeling of spin and
tidal effects. Binary mass-ratio is sampled
uniformly over the range $q\in[2,5]$ taking $q=\{2,3,4,5\}$ as our test points. The
mass of the NS is fixed to $\mns=1.35M_\odot$, and therefore knowing $q$ gives $\mbh$.
The spin of the NS is fixed to zero. The spin of the BH is sampled uniformly over the
range $\chibh\in[-0.75, +0.75]$, choosing $\chibh=\{-0.5, 0, +0.5, +0.75\}$ as our
test points. Finally, we vary the deformability parameter uniformly over 
$\lambdans\in[0,2000]$, choosing $\lambdans=\{500, 1000, 1500, 2000\}$ plus an additional
$\lambdans=800$ case. Waveform are generated for all possible combinations of these
parameters, and rescaled to different strengths with $\rho=\{20, 30, 50, 70\}$. For all
of the resulting signals, we perform a fully Bayesian parameter estimation aimed at
measuring binary masses and spins from the signal. Our analysis
templates differ from the signal in only one sense: all tidal terms are set to zero
in the former, i.e. they correspond to BHBH mergers.

In Fig.~\ref{fig:TN_chirpMassBias_vs_Lambda_SNR} we show the ratio of the systematic
bias in the measured binary chirp mass $\mchirp$ versus its true value, to the 
statistical uncertainty of the measurement itself. $\mchirp$ is the mass combination
that occurs at leading order in the perturbative expression for GW strain from 
compact binaries, and is therefore very precisely measurable. Even so, we find that
for $\rho=30$ the signal is not loud enough for the systematic bias in $\mchirp$
to exceed a fraction of the statistical uncertainty. When we increase the SNR to 
$\rho=50$ (middle column), we find that if we have two favorable conditions present:
(a) that the BH spin be $\gtrsim 0.4$, {\it and} (b) the NS's true deformability
be large enough that its $\lambdans \gtrsim 1000$, the systematic biases from
ignoring tidal effects in templates reach between $50-100\%$ of the statistical
uncertainty. Therefore at this signal strength, including terms involving the neutron
star's internal structure in waveform models becomes increasingly gainful. At 
even higher SNRs, say $\rho=70$ (right column), under the same favorable conditions
as listed above, the systematic bias in $\mchirp$ measurement {\it exceeds} its 
statistical measurement uncertainty. If the BH is even mildly spinning, say
$\chibh\gtrsim 0.2$, with the NS mildly deformable, say $\lambdans\gtrsim 800$, 
we still are affected in our $\mchirp$ measurement at the $50\%$ level by the 
use of point-particle waveform templates. Finally, we note that the measured (median)
$\mchirp$ is always higher than its true value. This is reasonable since the tidal 
deformation of NSs close to merger reduces the GW signal power in high frequencies, so
it would look like a BH-BH signal of a higher mass (and therefore lower merger or
peak GW-power frequency).

In the next figure, Fig.~\ref{fig:TN_EtaBias_vs_Lambda_SNR}, we show the same
information (i.e. the ratio of systematic bias to statistical uncertainty in parameter
measurement) for the symmetric mass-ratio $\eta$. Note that $\eta$
is bounded from above and below to within $\eta\in[0, 1/4]$. We immediately observe
that the measured (median) $\eta$ is always lowered below its true value due to the 
lack of $\lambdans$-dependent terms in the filter template model. This bias, as for 
the $\mchirp$ bias, decreases the merger frequency for the BH-BH template, which must
better match the NS-BH merger with its quietened merger and ringdown. At SNR $\rho=30$
the systematic bias in $\eta$ measurement stays mostly around $10-50\%$ of its statistical
measurement uncertainty. For the sole case of a very deformable NS with $\lambdans=2000$,
the missing tidal pieces in template models could cause errors that are comparable to 
that due to noise (i.e. statistical uncertainties). For a still stronger signal with 
$\rho=50$, we find that 
when the following conditions are fulfilled, the measurement of $\eta$ is {\it completely
compromised} due to template biases: (a) BH mass $\mbh\leq 5M_\odot$, (b) BH spin
$\chibh\gtrsim +0.4$, and (c) $\lambdans\gtrsim 1500$. Even under more moderate restrictions
on BH and NS parameters, we find that $\rho=50$ is loud enough to necessitate the 
use of tidal-modeled templates. This observation is further strengthened if we divert 
our attention to the third column in Fig.~\ref{fig:TN_EtaBias_vs_Lambda_SNR} that presents
the case for even higher SNRs ($\rho=70$).

Moving on from mass to spin parameters, we now consider the measurement of spin
angular momentum of the BH $\chibh$. The ratio of the systematic bias introduced into
and the statistical uncertainty associated with $\chibh$ measurement is shown in
Fig.~\ref{fig:TN_BHspinBias_vs_Lambda_SNR}. The presentation of information in this 
figure is identical to that in Fig.~\ref{fig:TN_chirpMassBias_vs_Lambda_SNR}
and~\ref{fig:TN_EtaBias_vs_Lambda_SNR}. A diverging colormap is used because both 
extremes of the colorbar range point to large systematic biases, while its zero (or
small) value lies in the middle. For the lowest considered SNR $\rho=30$, $\chibh$
bias is negligible, being below $50\%$ of its statistical measurement uncertainty.
For the most deformable NS structure {\it and} with a low-mass BH, the statistical
bias in $\chibh$ does become an important error. This is furthered at higher SNR, 
considering $\rho=50$ next (middle column of Fig.~\ref{fig:TN_BHspinBias_vs_Lambda_SNR}).
We find that for low-mass BHs (with masses below $4.5M_\odot$), the systematic 
measurement bias for BH spin exceeds its statistical uncertainty to become the 
dominant source of error. Not surprisingly, the systematic bias at both spin 
extremes pushes towards milder spins, since the prior excludes any more extreme
spins than $[-0.75, +0.75]$ and that pushes the bulk of the posterior inwards.
However as the signal's $\lambdans$ is increased from $500\rightarrow 2000$, the 
low-mass BH-NS binaries with large spins on the BHs show very large systematic biases
in spin values, that becomes larger than the statistical uncertinty for $\rho\geq30$
and reaches up to $4\times$ the same quantity as $\rho\rightarrow 70$.

Summarizing these results, we find that irrespective of system parameters, below a
signal-to-noise ratio of $30$, our measurements of mass and spin parameters of 
astrophysical BHNS binaries will remain limited by the intrinsic uncertainty due to 
instrument noise, and not depend on whether we include tidal effects in template
models. However, when the signal-to-noise ratio exceeds $30$ the systematic bias in
binary mass and spin measurements become comparable to and exceed the uncertainty
due to noise. Of the different parameters considered, we find that the measurement 
of $\eta$ degrades worst (in a relative-error sense) amongst all other parameters
of BHNS binaries, due to the use of BH-BH waveforms in deciphering an NSBH signal.

@@

\includegraphics[width=1.025\columnwidth]{plots/TTSNRThresholdFor100LambdaMeasurement_BHspin_BHmass_Lambda1500_0_CI90_0}
\includegraphics[width=1.025\columnwidth]{plots/TTSNRThresholdFor100LambdaMeasurement_BHspin_BHmass_Lambda2000_0_CI90_0}\\
